Reproducible development environments with rix

### 2023-12-13 paper.qmd
review of solutions to restore packages

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to list all the dependencies of the analysis
  clearly and ideally provide an easy way to install said dependencies.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as Docker, a containerization solution. This paper will present
  the Nix package manager, and an R package called {rix} that lowers
  Nix's learning curve for users of the R programming language.

@@

::: callout
The introduction is in principle "as usual". However, it should usually embed both the implemented *methods* and the *software* into the respective relevant literature. For the latter both competing and complementary software should be discussed (within the same software environment and beyond), bringing out relative (dis)advantages. All software mentioned should be properly `@cited`'d. (See also [Using BibTeX] for more details on {{< bibtex >}}.)

For writing about software JSS requires authors to use the markup `[]{.proglang}` (programming languages and large programmable systems), `[]{.pkg}` (software packages), back ticks like \`code\` for code (functions, commands, arguments, etc.).

If there is such markup in (sub)section titles (as above), a plain text version has to be provided in the {{< latex >}} command as well. Below we also illustrate how abbrevations should be introduced and citation commands can be employed. See the {{< latex >}} code for more details.
:::

@@

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then a build automation tool such as
[make]{.proglang} and ideally [Docker]{.proglang} should be added to the list.
[Docker]{.proglang} is a tool that makes it possible to run arbitrary code in a
completely controlled and isolated environment as a way to capture the complete
underlying system libraries.

>>>>

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to list all the dependencies of the analysis
  clearly and ideally provide an easy way to install said dependencies.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker].{proglang}, a containerization solution. This paper will present
  the Nix package manager, and an [R].{proglang} package called [rix]{.pkg} that lowers
  Nix's learning curve for users of the [R].{proglang} programming language.

@@

@@

Tools can be used at each of these steps to reach the gold standard of full
replication. Let's first consider the task of listing the software used.
[R]{.proglang} provides the `sessionInfo()` function whose output can be saved
into a file. Below is an example output of `sessionInfo()`:

```{r, eval = FALSE}
sessionInfo()
```

```
R version 4.3.2 (2023-10-31)
Platform: aarch64-unknown-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/aarch64-linux-gnu/openblas-pthread/libblas.so.3 
LAPACK: /usr/lib/aarch64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
 [9] LC_ADDRESS=C               LC_TELEPHONE=C            
[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C

time zone: Etc/UTC
tzcode source: system (glibc)

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods  
[7] base

other attached packages:
[1] nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163

loaded via a namespace (and not attached):
[1] compiler_4.3.2 Matrix_1.6-1.1 tools_4.3.2    splines_4.3.2 
[5] grid_4.3.2     lattice_0.21-9
```

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future) can read this file
and see which version of [R]{.proglang} was used, and which packages (and their
versions) were used as well. However, others would still need to install the
correct software themselves. An alternative to this is instead to use the [renv]{.pkg}
package which generates a so-called `renv.lock` file which also lists
[R]{.proglang} and package versions. Here is an example of such an `renv.lock` file:

```
{
"R": {
  "Version": "4.2.2",
  "Repositories": [
  {
   "Name": "CRAN",
   "URL": "https://packagemanager.rstudio.com/all/latest"
  }
  ]
},
"Packages": {
  "MASS": {
    "Package": "MASS",
    "Version": "7.3-58.1",
    "Source": "Repository",
    "Repository": "CRAN",
    "Hash": "762e1804143a332333c054759f89a706",
    "Requirements": []
  },
  "Matrix": {
    "Package": "Matrix",
    "Version": "1.5-1",
    "Source": "Repository",
    "Repository": "CRAN",
    "Hash": "539dc0c0c05636812f1080f473d2c177",
    "Requirements": [
      "lattice"
    ]

    ***and many more packages***
```

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific, library which doesn't interfere with the other,
main, library of the user. [renv]{.pkg} does not restore the [R]{.proglang}
itself though, so installing the right version of R needs to be handled
separately. Before continuing, it should be noted that other packages exist
which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
by @simonsohn2023 which makes it rather easy to install packages as they were
on CRAN at a given date. For example, the code snippet below install the
[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:

```
groundhog.library("
    library(purrr)
    library(ggplot2)",
    "2017-10-04",
    tolerate.R.version = "4.2.2")
```

These packages also get installed in a project-specfic library so there is no
interferences between these packages and other versions of the same packages
that one might use for other projects. Because [groundhog]{.pkg} does not
install [R]{.proglang} itself, users should either install the required version
themselves, or they should use the `tolerate.R.version` argument as shown in the
example above. Otherwise, [groundhog]{.pkg} would not continue with the
installation of the packages. Another such package, developed by @chan2023 is
[rang]{.pkg}, which also installs packages as they were on a given date.
Yet another way to install packages as they were on a give date is to use
the Posit Package Manager, which provides snapshots of CRAN. For example,
to install the required packages for an analysis as they were on the 30th of June
2023, one could add the following line to the `.Rprofile` file:

```
options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"))
```

The `.Rprofile` file is a file that gets read by R when starting a new session, which
means that every call to the `install.packages()` function will now install the packages
from this snapshotted mirror.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then a build automation tool such as
[make]{.proglang} and ideally [Docker]{.proglang} should be added to the list.
[Docker]{.proglang} is a tool that makes it possible to run arbitrary code in a
completely controlled and isolated environment as a way to capture the complete
underlying system libraries. Simply put, a `Dockerfile` provides a complete
description of the computational environment as well as the required instructions
to build the complete project. One can then use [Docker]{.proglang} to build
a so-called *image* which contains all the required software and instructions
to build the project. From this image, it is then possible to run a *container*
which actually runs the analysis.
